Predictive Models to Support Quoting of Fixed Fee Consulting Projects

### 2016-12-19 finalwriteup/Consulting Profitability Paper/Consulting_Profitability_Paper.tex
finished results and discussion of paper

Naive Bayes is a simple, fast method included in this work as it
provides a good baseline method against which to compare complex methods
(Caruana and Niculescu-Mizil 2006). It works by making conditional
independence assumptions about the explanatory variables in order to
simplify probability calculations for a categorical response variable.
For example, if predicting a binary variable, \(R\) (with classes 0 or
1) given two covariates \(X_1\) and \(X_2\), the equations for the
probability of both outcomes is as follows:

@@

The second research questions asks whether a predictive algorithm can
improve a company's profitability? This was analysed using a simple
decision-making scenario. The model with the highest AUC score may not
improve company profits the most, because this depends on \emph{which}
loss making and profitable projects are accurately being identified.
Therefore, all blended models as well as individual models were carried
forward for this analysis. A profit curve was built to compare methods,
which is a chart that plots the change in profit the company earns on
the y-axis vs.~the probability threshold on the x-axis. For this
decision making scenario, the threshold is the probability output value
(a value between 0 and 1 that the model predicts for each case), from
which it is decided whether a project will be accepted or rejected. For
example, if the output threshold is 0.6, and a model predicts a project
has probability of being a loss-making job of 0.7, the project is not
taken on by the case study company. All profits and losses from jobs
above the threshold were discounted. The y-axis of the profit curve is
the ratio of profit after rejecting projects based on model output to
historical profit:

@@

If the threshold were zero, all jobs were rejected, the profit would be
\$0, and the ratio to historical profits would be 0\%. If the threshold
was 1.0, all jobs were accepted and the profit would be the same as
historical profits, giving a ratio of 100\%. The aim was to find the
optimal threshold point where mostly loss-making projects were rejected,
resulting in higher profits. 100 profit curves were built for each
method to achieve a power of 0.8 and a 95\% confidence interval could
also be determined around each threshold point. The final expected
increase in profit combined with the percentage of projects to be
rejected presents a clear scenario that the case study business managers
could strategically assess for implementation.

@@

The predictive formula was re-structured so that the 11 explanatory
variables predicted a new response variable: profitable/unprofitable
projects. The five prediction methods, were initially built without
imputing data, by using the maximum amount of data possible depending on
the method. Boosted Trees and Naive Bayes are able to process data with
missing values, so all data could be input (Ridgeway 2015; Meyer et al.
2014). 100 models of each method were built. The violin plot below
summarises the AUC values produced by each method and the `violins' are
coloured according to whether the distributions significantly vary to
Boosted Trees using a critical value of 0.05.

\begin{figure}[htbp]
\centering
\includegraphics{Consulting_Profitability_Paper_files/figure-latex/reduced_violin-1.pdf}
\caption{Violin plot vertically illustrating the distribution of AUC
values from each of the methods when predicting profit/loss. Subsets of
the data were used for Logistic Regression and Random Forests in order
to provide datasets without missing values.}
\end{figure}

The AUC performance of Logistic Regression and the Random Forest
algorithms cannot be statistically differentiated from Boosted Trees.

Next, data imputation methods were trialed which would make the complete
data set available to Logistic Regression and Random Forests. Again, 100
models were required to achieve a power of 0.8 with respect to Boosted
Trees

\begin{figure}[htbp]
\centering
\includegraphics{Consulting_Profitability_Paper_files/figure-latex/mice_violin-1.pdf}
\caption{Violin plot vertically illustrating the distribution of AUC
values from each of the methods when predicting profit/loss. Each method
was fed the same imputed full dataset.}
\end{figure}

Boosted Trees, Logistic Regression, and Random Forest all performed
significantly better than the baseline algorithm, Naive Bayes, however
none outperformed Boosted Trees.

Testing Boosted Trees, Random Forests, Naive Bayes, Logistic Regression
and Bayesian Networks on the binary classification problem showed that
Boosted Trees, Logistic Regression and Random Forests performed best
(according to AUC). Naive Bayes and Logistic Regression were included as
baseline models and it was expected that the more complex models would
outperform these. Therefore it was surprising that results from 100
Logistic Regression models were not statistically significantly lower
than Boosted Trees.

A possible explanation for this, according to the literature, is that
there was not enough data for the ensemble trees to learn the complex
decision rules at which they excel. Trees tend to overfit the patterns
in a smaller training set. Logistic Regression on the other hand is
capable of only one decision boundary (which does not have to be
parallel to the variable axes) and is not prone to overfitting (Perlich,
Provost, and Simonoff 2003). This may explain Logistic Regression's
comparatively high performance on the case study's small but complex
data set.

Boosted trees, Random Forests and Logistic Regression all had mean AUC
values between 0.755 to 0.764. In summary, the individual binary
classification models performed well above random chance (AUC = 0.5).
Whether the model is worth implementing in the work place is dependent
on the extent to which the algorithm would improve `bottom line' profits
for the business and if the model can affect decisions in practice.

\begin{itemize}
\tightlist
\item
  NB and BN were bad - one sentence
\end{itemize}

@@

Blended models combine the predictions from each high-performing
individual model to create averaged or `blended' predictions. In this
case, the Random Forest, Boosted Trees, and Logistic Regression models
were chosen. Through variable importance studies, 6 variables were
included as metafeatures. All six methods were compared against the
original Logistic Regression model's AUC distribution (as it was not
statistically different from the ensemble trees)

\begin{figure}[htbp]
\centering
\includegraphics{Consulting_Profitability_Paper_files/figure-latex/blend_violin-1.pdf}
\caption{Violin plot vertically illustrating the distribution of AUC
values from each of the blending methods when predicting profit/loss.
100 models were built for each method.}
\end{figure}

The above plot shows that 4 methods had a higher mean AUC than the
original Logistic Regression model:

\begin{itemize}
\tightlist
\item
  the simple blended Logistic Regression
\item
  the simple average
\item
  and the two Boosted Tree models.
\end{itemize}

The simple averaged method and simple logistic regression could achieve
a statistical power of 0.8 with 150 samples (models), whereas the
boosted trees required thousands. The additional 50 models were run for
the simple logistic regression and simple average to reveal similar
distributions that were indeed statistically significantly higher than
the individual logistic regression AUC distribution. P-values from two
sample t-test were 0.0019 and 0.0021 respectively.

Blended models improved the mean AUC from 0.759 to 0.77, which is an
increase of 1.4\%. It was expected that model blending would improve
predictions because it combined the strengths of individual models with
different theoretical foundations.

The trials of different blending methods demonstrated that again, the
simplest methods worked best with the case study data. In this incident,
averaging the results of the best three single models (Logistic
Regression, Random Forests, and Boosted Trees) or taking a Logistic
Regression of the three models outperformed more complex Boosted Trees,
Random Forests, and Logistic Regression blends that facilitated
interaction of the individual model results and original variables
(meta-features). Simple blended models could perform better than complex
methods if the data is not big enough for the complex models to learn
the more intricate patterns at which they excel. This was observed in
the single model methods. their success in the 2009 Netflix competition
generated some publications. The Netflix data set comprised of almost
3,000,000 observations, so the size of the data could have enabled the
success of complex blending methods such as FWLS (Sill et al. 2009)

@@

This chapter first presents the full range of method results in terms of
improvements to the case study's bottom line. A business decision making
scenario was created so that profit curves based on the decision rule
could be built and analysed. From the profit curves, optimal probability
thresholds could be derived for each method (individual and blended
methods). The predictive models output a `probability' between 0 and 1
that each project will be a loss making job (where probability = 1
indicates a loss making job). The question the arises, at what
probability would a decision-maker round the probability up to 1 or down
to 0? And what business decision would then be made? To find the
threshold point for rounding, an experimental business-scenario was
tested. At a given threshold, all projects with probability outputs
above the threshold were considered too risky, and were rejected. All
profits and losses from these projects were forfeited, while the profits
from the remaining jobs (below the threshold) were summed to give a
revised total profit. This profit calculation was made on a range of
thresholds between 0 and 1 at 0.05 increments. The total profits were
plotted for each threshold and joined to make a profit curve

The plot below illustrates the distribution of profit curves for each
method, where the solid lines join the mean values at each threshold
point. The grey ribbon illustrates the 95\% confidence interval for the
profit improvement ratio at each threshold point for the 100 models.

\begin{figure}[htbp]
\centering
\includegraphics{Consulting_Profitability_Paper_files/figure-latex/profit_curve-1.pdf}
\caption{Profit curves summarising results from 100 models of 9 methods:
3 simple blends, 3 complex blends, and the original 3 best methods.}
\end{figure}

Two blended models clearly outperformed the individual models as shown
in the above plot with higher profit ratios as well as tighter
confidence intervals. The simple Logistic Regression blend performed
best with the highest mean profit ratio of 109\% with a standard
deviation of 1.28\%. \%. This means that for the simple blended Logistic
Regression model, if all jobs above the probability threshold 0.6 were
rejected, the profits would increase on average by 9\% in comparison to
historical profits.

Profit curves from the simple average blend and Logistic Regression
blended models outperformed the complex blended models, which follows
logically from their significantly higher AUC distributions. The simple
average blended model produced a profit curve with an almost identical
profit improvement (and standard deviation) to the Logistic Regression
blend. In the data, projects assigned a probability higher than 0.6
represent only 4.3\% of all projects.

The shaded grey 95\% confidence intervals in the Profit Curve plot shows
that with the 100 sample models that were analysed, none of the lower
bounds for the original methods' or complex blending methods were above
100\% on the y-axis. That is, it cannot be said with 95\% confidence
that the original methods would produce a mean profit higher than
historical profits in the given decision framework. Clearly this level
of certainty is not beneficial for the case-study business to adopt.

It is not clear why simple linear models and the averaging model
outperformed ensemble tree blending methods. As previously stated, the
ensemble trees may not have received enough data to adequately learn the
complex series of rules they develop. The only differences between FWLS
and the simple Logistic Regression (that performed best) were two
additional explanatory variables and four interaction terms. The number
of variables was not high for the amount of data since according to
Peduzzi et al. (1996), 10 events per explanatory variable or more avoids
the risk of biased estimation of variable coefficients in Logistic
Regression. The data contained 315 events per explanatory variable and
did not pose that risk. Nevertheless, the additional variables must have
added misleading noise to the model.

To conclude, because of the promising AUC results, it was logical that
the models translated into financial benefits for the company. The 9\%
improvement in profits produced by the simple blended Logistic
Regression is reasonable, and may be high enough to trigger further
cost-benefit analyses and the development of a more comprehensive
framework describing different decision scenarios.

\begin{itemize}
\item
  decision scenario
\item
  profit curves
\item
  prototype decision support interface
\end{itemize}

@@

The general aim was to use statistical techniques to predict the
profitability of projects for a case study consulting business using
their internal CRM data. This was rigorously completed by approaching
the prediction of `profitability' as a binary classification problem
predicting either profit or loss. Several statistical and machine
learning approaches were applied including Naive Bayes, Bayesian
Networks, Linear Regression, Random Forests, gradient Boosted Trees as
well as multiple methods of blending the individual models' output.

AUC = 0.76). This was achieved by 3 individual methods while various
techniques that blended the individual methods improved results further
(AUC = 0.77).

predictive models could be shown to improve the overall profitability
(bottom line) of the case study business. A range of probability
threshold values were trialed for each method using a decision framework
where projects scored below the threshold were accepted, while projects
above the threshold were rejected. If a project was rejected, the
profits and losses were forfeited, and the remaining accepted profits
and losses were summed.

Final results showed the simple Logistic Regression blend of the
individual Logistic Regression, Random Forest and Boosted Tree models
improved profits the most. The 95\% confidence interval for these
improvements was between 6.5\% and 11.5\% using a probability threshold
of 0.6 (approximately 4.3\% of projects).

These results contribute significantly to the research in cost
estimation in three ways: the applied methods, the decision framework,
and the appeal to user trust. Ensemble tree methods and blending had
been applied minimally to cost estimation previously, even though
ensemble trees provide insight into model structure while predicting at
a similar level to Neural Networks. Next, previous studies have verified
predictive accuracy but stopped short of how the algorithm would affect
decisions and what the measured benefits would be. This study presented
a clear framework for how a business could improve profits by applying
the algorithm.

Further work is required to test user confidence in the output. Another
topic identified for future research was to test how well managers
estimate some numeric project input variables. In particular, time span,
team size, total invoiced amount, and percentage of hours completed by
professionals should be tested as these variables were calculated post
project completion. Time span and total invoiced amount were discretised
into wide categories, which should be easier for a manager to choose
between.

future work: user interface.

Overall, this work has successfully built a mathematical blend of
Logistic Regression, Random Forests, and Boosted Tree models, from a
consulting company's internal project data. This blended model can
predict whether a project will be profitable or not and in a reasonable
decision framework, can guide managers in rejecting financially risky
projects and improving profitability of the business.

@@

@@

\hypertarget{ref-Hastie2009}{}
Hastie, Trevor, Robert Tibshirani, and Jerome Friedman. 2009. \emph{The
Elements of Statistical Learning: Data Mining, Inference, and
Prediction, Second Edition}. Book. Springer New York.
\url{http://qut.summon.serialssolutions.com/2.0.0/link/0/eLvHCXMwzV1bS8MwFA7iQBwD3dR6mRAQfNk6trb0Ivi0TQY6EJwgQxi9JLKHddpVf7_nJE269cFnH9NCm_Y7JF--cyPEtnp9s7Im8JCzAWxVjFluEnMWBF4MyyfzE4-HVsR29Tfd9rO89s-BrzBiPSwRH8tocBGfgTxSlGFGrbpQQFAAGIV52JmKxhByhShS_lQg53OGzhsR9IHSKJ6YsfDxcttlP8vYzzrrTEJ4vjYNGaUN8EcoeYfpUofnMCyKACQZ-O6qMEelNAQVpUEpjR1VUGrnKIpucN_BcvnlzqK86ZUNR4cBwrbpes4tFjlf4Yfds9R8fYH9dAA8FGnufK71MnR_ArkqHO7iTW5RQEm_uU4aMEPxsA9subLFFGbHpMYwfaRJ9ljaIkeqaQYt1tAWqU91odzNCXkH0KgCja453QKNKtDuKEJGJWRdqgHrUoCLlnB1qQSLFmCdkpuH8Ww4MXemu_iUBUYW8r_YZ6QRYppDmot0yMQgNQ6myQzkCQZ8l0EO3oKnkT95HMphUw17G5Gz1_vKDaAlwrJNt-edE2pb3PEiL-aRkzh-AKfFqO8nPse2BbFt9S9I-69JXf59-4oclpbTJvt59s2uRQ7tL6t_R8w}.

@@

\hypertarget{ref-Lovallo2003}{}
Lovallo, Dan, and Daniel Kahneman. 2003. ``Delusions of Success: How
Optimism Undermines Executives' Decisions.'' Generic. HARVARD BUSINESS
SCHOOL PUBLISHING CORPORATION.
\url{http://qut.summon.serialssolutions.com/2.0.0/link/0/eLvHCXMw3V3Pb9MwFLYQQxMSEgxYKAPJBwQHlCqO61-TdoC2aNImcdgmcYscxwGktoMlQfvzeXYSJy1M4swxjZuq_uz3vs95PxCi6TSJd2yCAU-T53pmjbIgoblMCqGYyEtpEpoWs-3zt9Bqb_jsfwB-YVdN1Ye3XTS-IaLT_a553GcwEGvXFsO3O1q7kPf3y1trGmf0fDzFouu5U41Za99AKETJb79ROL_-pbtXOIthsZ3pbxvbna-2iexbJww0RKMGq9l5srHVlGS0OsTIBDI-cqakTczcrnO9439CVCCIZQ6ml751Vc_XxXdTn9hNfHUBDhZ4q1cFH5d3utk_feqIVDxwOT5N9Vd24ZnE5RP0uC_pjT-02B2ge3bzFO33k_sMzQOE-LrEHYTHGADEPYB4ABAPAL7DAb7n6OrT8nJ-GnfNLuKvBChkzFTJcgnkVadGg1XkVllKiXYEDHYLYxKUK3imPCfGklIwA8KxYJQpSsAaSHqIHmmXFLGpffJkEaG9EhayjRyriOBfRmj_izpfyNOzeXt50F9OK5_hN_1ZRzC7fh_EfCpeICxEIg2fcebqtoHtzlOic26UYobKWZFMUOQmPXObor7RJiMJT0BSJGqCDnscsmK1yqgrlQfklMBXWlSyH209lQwoEgPdDnfetDCFO2lWpRmoVBDJqRBOq9a3NTxgZ1i_cOABY3jDfSdkYIhgzEueCSL_MmzeFcl3xSHql3f-6BF6OGyaV-h-fdPY1z5t-DdLboxJ}.

@@

\hypertarget{ref-e1071}{}
Meyer, David, Evgenia Dimitriadou, Kurt Hornik, Andreas Weingessel, and
Friedrich Leisch. 2014. \emph{E1071: Misc Functions of the Department of
Statistics (E1071), Tu Wien}.
\url{https://CRAN.R-project.org/package=e1071}.

@@

\hypertarget{ref-Pai2013}{}
Pai, Dinesh R., Kevin S. McFall, and Girish H. Subramanian. 2013.
``Software Effort Estimation Using a Neural Network Ensemble.'' Journal
Article. \emph{Journal of Computer Information Systems} 53 (4): 49--58.
\url{http://qut.summon.serialssolutions.com/2.0.0/link/0/eLvHCXMwnV1Nb9QwEB0VekFClPLVAJV8gGPAH4kbc4GlyraIbhftpurRysTe06os3VTqz8fjTbawBanqKZHiOIntvBmPPe8BKPmBpxuYgMHsmgALqtHauDrYJRTaZNlB4b3XiBvxty99pGDV2z1IRuR2PxsKmn8UAUm1IcKwz4tfKclI0XJrp6nxALaJVoSG_Ej-WC8rFHnUE4x_Fmlv_e1WUh7I1fKfFiham-EO9Nuf-tjX_FZywW02x3t_xFN40nmlbLAaRruw5S-ewU6v-MA6AHgOn6bjYXU-mJQs-LfjScXKafVtFMNcjAQ8jtiAnZZnk8FJOFTn48l3Vp5Oy9HXk_IFnA3L6vA47dQX0gWlFaWNwjwjOjfDZ43hssFaovNCu7wIZwpnJrhmuSPtK2xE0QhZ68IJn3PnMzxQL-FxTbv0L9qYzef2gKkw-eJZ1BnFzGcKQ_Ua3SxUyRELTGC_byZbI8V6mnZpbxopgVfr624-t2G-QynmSooE3q-6zy5WLB1W2qW03BYBoSTdrbltr9sE9jbKkaChKLgyCbz7s-PXBWjaGJ6guInGPQFxl2KHHcU6UQu0r__73m_gkYwCG7QB-C08bC-v_H5MOv0NwrD2yw}.

\hypertarget{ref-Peduzzi1996}{}
Peduzzi, Peter, John Concato, Elizabeth Kemper, Theodore R Holford, and
Alvan R Feinstein. 1996. ``A Simulation Study of the Number of Events
Per Variable in Logistic Regression Analysis.'' Journal Article.
\emph{Journal of Clinical Epidemiology} 49 (12): 1373--9.

@@

\hypertarget{ref-Provost2013}{}
Provost, F., and T. Fawcett. 2013. \emph{Data Science for Business}.
Book. O'Reilly. \url{https://books.google.com.au/books?id=_1b4nAEACAAJ}.

\hypertarget{ref-gbm}{}
Ridgeway, Greg. 2015. \emph{Gbm: Generalized Boosted Regression Models}.
\url{https://CRAN.R-project.org/package=gbm}.

>>>>

Naive Bayes is a simple, fast method included in this work because,
similarly to Logistic Regression, it provides a good baseline method
against which to compare complex methods (Caruana and Niculescu-Mizil
2006). It works by making conditional independence assumptions about the
explanatory variables in order to simplify probability calculations for
a categorical response variable. For example, if predicting a binary
variable, \(R\) (with classes 0 or 1) given two covariates \(X_1\) and
\(X_2\), the equations for the probability of both outcomes is as
follows:

@@

The second research question asks whether a predictive algorithm can
improve a company's profitability. The model with the highest AUC score
may not improve company profits the most, because the magnitude of the
profits and losses from each project have not been analysed. Therefore,
all blended and individual models were carried forward to be assessed in
a value framework.

Once a model outputs scores for the probability of each job being
loss-making, the question then arises at what probability would a
manager round the probability up to 1? And then what action would be
taken? A threshold point between 0 and 1 can dictate when a project is
rounded up, and classified as `loss-making'. For this analysis, a simple
decision scenario is tested where a manager rejects all projects with
output above the threshold. For example, if the threshold is 0.6, and a
project is predicted to have a probability of loss-making of 0.7, the
project is rejected and profits or losses associated with the project
are ignored.

To test each threshold between 0 and 1, profit curves were built for
each method. This chart plots the probability threshold on the x-axis
and the ratio of model-influenced profits over historical company
profits on the y-axis. The ratio represented by the y-axis is calculated
using the formula below:

@@

If the threshold were zero, all jobs were rejected, the profit would be
\$0, and the ratio to historical profits would be 0\%. If the threshold
was 1.0, all jobs were accepted and the profit would be the same as
historical profits, giving a ratio of 100\%. The total profits were
plotted for each threshold and joined to make a profit curve. The aim
was to find the optimal threshold point where mostly loss-making
projects were rejected, resulting in higher profits. 100 profit curves
were built for each method to achieve a power of 0.8 and a 95\%
confidence interval could also be determined around each threshold
point. The final expected increase in profit combined with the
percentage of projects to be rejected presents a clear scenario that the
case study business managers could strategically assess for
implementation.

@@

The violin plot below summarises the distributions of the 100 AUC values
produced by each method. The `violins' are coloured according to whether
the distributions significantly vary to Boosted Trees using a critical
value of 0.05. Boosted Trees was chosen as the comparison method because
it was expected to perform well.

\includegraphics{Consulting_Profitability_Paper_files/figure-latex/mice_violin-1.pdf}

The AUC distrubtions of Logistic Regression and the Random Forest
algorithms could not be statistically differentiated from Boosted Trees,
while the Naive Bayes distribution was significantly lower. The three
high-performing methods had AUC values between 0.755 and 0.764. This
indicates predictions were well above random chance, where AUC = 0.5.

Naive Bayes and Logistic Regression were included as baseline models and
it was expected that the more complex models would outperform these.
Therefore it was surprising that results from 100 Logistic Regression
models were not statistically significantly lower than Boosted Trees. A
possible explanation for this, according to the literature, is that
there was not enough data for the ensemble trees to learn the complex
decision rules at which they excel. Trees tend to overfit the patterns
in a smaller training set. Logistic Regression on the other hand is
capable of only one decision boundary (which does not have to be
parallel to the variable axes) and is not prone to overfitting (Perlich,
Provost, and Simonoff 2003). This may explain Logistic Regression's
comparatively high performance on the case study's data set, which is
small for the range and complexity of the described projects.

@@

Six blended models combined the predictions from the three
high-performing individual models: Logistic Regression, Random Forests,
and Boosted Trees to create `blended' predictions. There were 3 simple
blended models, and 3 complex models that interacted original covariates
with the individual model predictions. Through variable importance
analysis on the original covariates and the 3 new variables containing
model predictions, only 6 original covariates were included as
metafeatures. All six blended methods' AUC values were compared against
the Logistic Regression AUC distribution, which was chosen because it
was the simplest high performing individual model.

\includegraphics{Consulting_Profitability_Paper_files/figure-latex/blend_violin-1.pdf}

None of the blended methods were significantly higher than the
individual Logistic Regression model with data from 100 models. However,
the simple averaged method and simple logistic regression could achieve
a statistical power of 0.8 with 150 samples (models), whereas the other
blended methods required thousands. 50 more models were run for the
simple logistic regression and simple average blend to confirm their AUC
distributions were significantly higher than the individual Logistic
Regression. P-values from two sample t-test were 0.0019 and 0.0021
respectively.

Blended models improved the mean AUC from 0.759 to 0.77, which is an
increase of 1.4\%. This confirms the expectation that model blending
would improve predictions by combining the strengths of different
methods' theoretical foundations.

Again, the simplest methods worked best. In this case, averaging the
results of the best three individual models or taking a Logistic
Regression of the three models' outputs outperformed the complex Boosted
Trees, Random Forests, and Logistic Regression blends that facilitated
meta-feature interactions. Simiarly to the individual models, Simple
methods may outperform complex methods if the data is not big enough for
complex methods to learn the intricate patterns at which they excel. The
Netflix data set comprised of almost 3,000,000 observations, so the size
of the data could have enabled the success of their complex blending
methods (Feature Weighted Linear Stacking) (Sill et al. 2009)

The outcomes in this section show that statistical and machine learning
methods can predict whether a consulting project will be profitable or
not to some extent. The average, or weighted average (logistic
regression) blend of individual Logistic Regression, Boosted Trees, and
Random Forests models produced the highest AUC values. However, the
business value of the model has not been proven. The following section
explores whether the models would improve overall profits for the case
study business.

@@

This chapter transports model results into the value framework to assess
impact on the case study's overall profitability. The plot below
illustrates the distribution of profit curves for the 100 models built
using each method. Solid lines join the mean values at each threshold
point and the grey ribbon represents the 95\% confidence interval for
the profit ratio.

\includegraphics{Consulting_Profitability_Paper_files/figure-latex/profit_curve-1.pdf}

From the profit curves, the optimal probability thresholds are where the
profit curves peak, and peaks above 100\% indicate the model assisted
profits outperformed historical profits. The two simple blended models
clearly outperformed all other models with higher profit ratios as well
as tighter confidence intervals. The simple Logistic Regression blend
performed best with the highest mean profit ratio of 109\% and a
standard deviation of 1.28\%. This means that if all jobs above the
probability threshold 0.6 were rejected, the profits would increase on
average by 9\% in comparison to historical profits. Only 4.3\% of
projects were rejected in this case.

Profit curves from the simple blended models outperformed the complex
blended models, which follows logically from their significantly higher
AUC distributions. The shaded grey 95\% confidence intervals show that
none of the lower bounds for the original methods' or complex blending
methods were above 100\% on the y-axis. That is, it cannot be said with
95\% confidence that these methods would produce a mean profit higher
than historical profits in the given decision framework. It is
interesting that despite the reasonable AUC values, the profit curves
for the individual and complex blended models do not make a complelling
case for adoption into business practice.

The simple blended Logistic Regression model on the other hand, a mean
9\% improvement in profits, with a lower bound of 6.5\%. In terms of
implementing the model, this improvement may be high enough to trigger
further development and analysis of a more sophisticated framework
describing decision making scenarios.

@@

Complex ensemble tree methods did not outpereform simple linear methods,
however, the simple blended predictions of all methods

The general aim of this work was twofold. First, statistical and machine
learning techniques were tested to determine whether the profitability
of projects for a case study consulting business could be predicted
using their internal CRM data. This was rigorously completed by
approaching the prediction of `profitability' as a binary classification
problem with classes profititable or loss-making. Naive Bayes, Logistic
Regression, Random Forests, gradient Boosted Trees as well as simple and
complex blending methods were applied to the problem. Secondly, this
project tested whether predictions from the models could be proven to
improve the overall profitability of the case study company and by how
much.

Predicting whether a job would be profitable or not (binary
classification) was possible to a level significantly better than random
assignment (AUC = 0.76). This mean statistic was achieved by 100
different training/testing divisions of the data, built for each of the
3 individual methods. Simple averaging and simple Logistic Regression
blending methods improved results further to a mean of AUC = 0.77.

Two models could also be shown to improve the overall profitability of
the case study business. This was done by converting the binary
predictions from the first analysis into profit curves that plotted
model-influenced profit improvements over a range of probability
threshold values. To determine model-influenced profit improvements, a
managerial decision framework was used where projects scored below the
threshold were accepted, while projects above the threshold were
rejected. If a project was rejected, the profits and losses were
forfeited, and the remaining accepted profits and losses were summed.

100 training/testing divisions of the data were user to produce 100
profit curves so that a 95\% confidence interval could be framed around
the curves. Final results showed the simple Logistic Regression blend of
the individual Logistic Regression, Random Forest and Boosted Tree
models improved profits the most. The 95\% confidence interval for these
improvements was between 6.5\% and 11.5\% using a probability threshold
of 0.6 (approximately 4.3\% of projects were rejected).

These results contribute significantly to the research in cost
estimation in two ways: the applied methods, and the value assessment
based on a managerial decision framework. Ensemble tree methods and
blending had been applied minimally to cost estimation previously, even
though ensemble trees provide insight into model structure while
predicting at a similar level to Neural Networks. Next, previous studies
compared accuracy of their predictive models, but stopped short of
computing how the algorithm would affect decisions and what the measured
benefits would be. This study presented a clear framework for how a
business could improve profits by applying the algorithm.

Further work is required to measure user confidence in the output.
Another topic identified for future research is to test how well
managers estimate input variables at project conception. This is
paraticularly relevant for time span and total invoiced amount which
were discretised into 4 to 5 categories.

Overall, this work has successfully built a mathematical blend of
Logistic Regression, Random Forests, and Boosted Tree models, from a
consulting company's internal project data. This blended model predicts
whether a project will be profitable or not and in a reasonable decision
framework, can guide managers in rejecting financially risky projects
and improving profitability of the business by a mean of 9\%.

@@

\section{Figure Captions}\label{figure-captions}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  mice\_violin: Violin plot vertically illustrating the distribution of
  AUC values from each of the methods when predicting profit/loss. Each
  method was fed the same imputed full dataset.
\item
  blend\_violin: Violin plot vertically illustrating the distribution of
  AUC values from each of the blending methods when predicting
  profit/loss. 100 models were built for each method.
\item
  profit\_curve: Profit curves summarising results from 100 models of 9
  methods: 3 simple blends, 3 complex blends, and the original 3 best
  methods.
\end{enumerate}

@@

\hypertarget{ref-Hastie2009}{}
Hastie, Trevor, Robert Tibshirani, and Jerome Friedman. 2009. \emph{The
Elements of Statistical Learning: Data Mining, Inference, and
Prediction, Second Edition}. Book. Springer New York.

@@

\hypertarget{ref-Lovallo2003}{}
Lovallo, Dan, and Daniel Kahneman. 2003. ``Delusions of Success: How
Optimism Undermines Executives' Decisions.'' Generic. HARVARD BUSINESS
SCHOOL PUBLISHING CORPORATION.

@@

@@

\hypertarget{ref-Pai2013}{}
Pai, Dinesh R., Kevin S. McFall, and Girish H. Subramanian. 2013.
``Software Effort Estimation Using a Neural Network Ensemble.'' Journal
Article. \emph{Journal of Computer Information Systems} 53 (4): 49--58.

@@

\hypertarget{ref-Provost2013}{}
Provost, F., and T. Fawcett. 2013. \emph{Data Science for Business}.
Book. O'Reilly.
